RCloud: Integrating Exploratory Visualization, Analysis and Deployment

### 2014-03-28 introduction.tex
update

% What's the area
Consider the emerging role of a data science team within an
organization today \cite{Keim:2008:VAS}.

% How do people do it today?
Individual data scientists and statisticians usually work on loosely
related problems, and must devise effective ways to share their findings
and move results from exploratory data analysis to automated diagnostics
and reports that are deployed for wider consumption.
They generally rely on a patchwork of resources shared informally,
or at best, through systems that attack individual tasks or functions
in the data analysis workcycle. % might cite information visualization reference model or pirolli/card

@@

Second, EDA environments often assume a single-developer perspective,
while data scientist teams could greatly benefit from convenient searching
and sharing of scripts, data sets, data feeds, experiments, annotations,
and automated analyses and recommendations. These capabilities have had
tremendous impact on information retrieval in general but are well
beyond what existing visual analytics environments provide, or what
is found in today's repositories and version control systems.

% What are we going to do about this?
\todo{Sell idea, not system} In this paper, we investigate the hypothesis
that the practice of visual analytics can advance by adopting techniques
from information retrieval and collaborative work environments. To investigate
these ideas, we designed and deployed a prototype environment called RCloud,
that supports collaborative data analysis, visualization and web deployment.
We will report on the design decisions, tradeoffs and limitations, and compare
the concepts in RCloud to similar proposals.

\todo{Define the requirements for the system}

We begin by noting the main requirements for a data analysis environment for
collaboration and transfer of results into production.

Main requirement is to satisfy these two sets of users: R coders and
managers. They work differently and want different things. What should
the infrastructure look like so they both coexist in peace? If this is
such a fundamental problem, why have other tools not addressed it?
Discuss Shiny (maybe in the disc. section).

\begin{itemize}
\item Constant technology transfer. Experiments in notebooks developed
by analysts need to become accessible in the production environment,
making it easy to push new features, prototypes, etc.
\item Dashboarding. To support the above, there should be a mode in which non-experts
can operate the experiments without coding.
\item Version control as a primary concept in the repository. There
needs to be a way to control the visibility of prototypes as they are
being developed and tested.
\item Coexistence in the current data ecosystem. We perceive the
isolation of current exploratory visualization and data analysis
environments to be a serious threat to a wider adoption of modern
techniques.
\item Modern interactive technologies. current isolation
notwithstanding, we find there is business demand for interactive
visualization tools. The main challenge here is how to integrate
these technologies so data scientists and statisticians can target
them, \emph{and} business-facing users can use them.
%% Scalable technologies like crossfilter exist. Want to support those in the context of an R environment. Need to support full 2-way communication between computing in the cloud and interaction on the web.
%% \item Other requirements e.g. Kandel's user study. Not our contribution but things we must address, mentioned in related work section. Heer/Agrawala design considerations for collaborative visual analytics.
\end{itemize}

% What is the scope of our solution?
Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
for statistical computing and graphics. R is one of the most popular
systems for data analysis and is well-supported by
a complex ecosystem of compatible packages, software
tools and other resources.
%
Of importance to us, there are existing
packages for high performance computing in the cloud,
and for generating web graphics. % no, there aren't. we wrote those!

>>>>

% What's the area
Consider the emerging role of a data science team within an
organization today \cite{Keim:2008:VAS}.
%
% How do people do it today?
Individual data scientists and statisticians usually work on loosely
related problems, and must devise effective ways to share their findings
and move results from exploratory data analysis to automated diagnostics
and reports that are deployed for wider consumption.
They generally rely on a patchwork of resources shared informally,
or at best, through systems that attack individual tasks or functions
in the data analysis workcycle. % might cite information visualization reference model or pirolli/card

@@

Second, EDA environments often assume a single-developer perspective,
while data scientist teams could greatly benefit from convenient searching
and sharing of scripts, data sets, data feeds, experiments, annotations,
and automated analyses and recommendations. These capabilities have had
tremendous impact on information retrieval in general but are well
beyond what existing visual analytics environments provide, or what
is found in today's repositories and version control
systems. \todo{add cites to reco system work}

In addition, take the recent trend in interactive visualization
technologies. Let us consider prominent designers and developers of
open source software, Jeffrey Heer, and Hadley Wickham. Heer has
written (or helped write) Prefuse, Flare, Protovis, d3, and Vega, in
that order. All but the first of these libraries target web
technologies, and all but the second use open standards. Wickham has
written ggplot, ggplot2, and ggvis. ggplot and
ggplot2 generate static graphics, while ggvis uses Shiny and Vega to
provide easy interactive visualizations in R, by opening a separate
web server for each call. \todo{Add cites for all of these}

% What are we going to do about this?

%% \todo{Define the requirements for the system}
%% We begin by noting the main requirements for a data analysis
%% environment for collaboration and transfer of results into production.

%% Main requirement is to satisfy these two sets of users: R coders and
%% managers. They work differently and want different things. What should
%% the infrastructure look like so they both coexist in peace? If this is
%% such a fundamental problem, why have other tools not addressed it?
%% Discuss Shiny (maybe in the disc. section).

We see the present state-of-the-art in environments for data science
teams as tremendously powerful, but ultimately fragmented. Rather than
design a completely new solution that risks even further
fragmentation, we wish to find and remedy the shortcomings of data
science as we see it practiced today. The following requirements
summarize what we consider to be the main pain points:

\begin{enumerate}

\item Technology transfer. In our experience, development
of new analyses and visualizations is done by \emph{hackers}
and \emph{scripters} (where here are use Kandel et al.'s
terminology~\cite{Kandel:2012:EDA}) who are, by definition, familiar
to some extent with programming and software development.
On the other hand, \emph{application
users} are the ones extracting value from the data using the tools
developed by hackers and scripters. Traditionally, the communication
between these two world is done by IT staff. In an environment where
business needs change possibly many times a week, manual IT support
does not scale: there needs to exist
infrastructure to connect the two worlds.
%% The two classes of users
%% have quite different perspectives but we
%% infrastructure for continuous deployment of visualizations.

\item Coexistence. In the current data science ecosystem, the
isolation of current exploratory visualization and data analysis
environments is a threat to wider adoption of modern techniques for
both sides. We specifically see the richness of interactive
visualization tools isolated from the richness of statistical
programming environments.
We wish to provide infrastructure such that
developments on one side can more easily be employed on the other. In
addition, this system itself should be easily accessible
to the rest of the organization.

\item Discoverability. From our experience in working with data
analyst teams, we find that repeated work (``how can I connect to
database X?'', ``how do you get clean data from column Y from feed
Z?'') typically arises from lack of communication of solved
subproblems. We wish to provide transparency about the work of
team members in the organization.

\end{enumerate}

Technologies which fail to address these requirements
are holding back the practice and impact of interactive visual
analysis. In this paper, we discuss the design and deployment of
RCloud, a prototype environment that supports collaborative data
analysis, visualization and web deployment satisfying all requirements
above. We will report on the design decisions, tradeoffs and
limitations, and compare the concepts in RCloud to similar proposals.

RCloud uses the R language~\cite{RCoreTeam:2013:R} as the main provider of
statistical analysis tools. In fact, as we describe in
Section~\todo{where?}, RCloud notebooks are sequences of R snippets
interspersed with Markdown for easy generation of reports. In
Section~\todo{where?}, we describe how RCloud enables
interactive visualization widgets in this setting. In
Section~\ref{sec:limitations}, we discuss some consequences of
designing for a single language, and how this was received by
our (small, but active) user community.

%% Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
%% for statistical computing and graphics. R is one of the most popular
%% systems for data analysis and is well-supported by
%% a complex ecosystem of compatible packages, software
%% tools and other resources.
%
%% Of importance to us, there are existing
%% packages for high performance computing in the cloud,
%% and for generating web graphics. % no, there aren't. we wrote those!

The facilities for \emph{sharing, viewing and embedding} RCloud
notebooks address technology transfer, and are described in
Section~\todo{where?}. RCloud notebooks are edited and shared as web
pages. In order to support the tight integration of R and HTML5
technologies, RCloud uses a full, bi-directional communication channel
between the web browser and a running R process, described in
Section~\todo{where?}.

Because RCloud embraces technologies that are part of the computing
environment (specifically, inter-process communication via HTTP and
web services as remote procedure calls), we find it easier to
integrate RCloud with satellite efforts in corporate environments. We
discuss this architecture and consequences in
Section~\ref{sec:system}.

The requirement of continuous technology transfer creates challenges
about the \emph{stability} of the many moving pieces: if notebooks are
always updated and deployed automatically, how can the system handle
changing requirements and public interfaces? We discuss our choice of
transparent, integrated version control for RCloud notebooks in 
Section~\ref{sec:notebooks}.

%% Dashboards are a solution to requirement 1, using solutions of
%% requirement 2.
%% Requirement 2: \emph{how is accessibility supposed to happen?}

%% Version control as a primary concept in the environment is a
%% consequence of requirement 1. There
%% needs to be a way to control the visibility of prototypes as they are
%% being developed and tested.

%% Tight integration between R process and web browser is a solution to
%% requirement 2

%% Current isolation notwithstanding, we find there is demand for interactive
%% visualization tools. We do not believe that moving the \

%% The main challenge here is how to integrate these
%% technologies so data scientists and statisticians can target
%% them, \emph{and} business-facing users can use them.
%% Scalable technologies like crossfilter exist. Want to support those in the context of an R environment. Need to support full 2-way communication between computing in the cloud and interaction on the web.
%% \item Other requirements e.g. Kandel's user study. Not our contribution but things we must address, mentioned in related work section. Heer/Agrawala design considerations for collaborative visual analytics.

% What is the scope of our solution?

### 2014-03-28 paper.tex
merge
update

% \title{RCloud: Integrating Exploratory Visualization, Analysis and Deployment}
\title{RCloud: DevOps for Visual Analysis and Data Science}

@@

%% Abstract section.
\abstract{ Consider the emerging role of a data science team within an
  organization. Individual data scientists and statisticians usually
  work on loosely related problems, and must devise effective ways to 
  share their findings and move results from exploratory data analysis
  into automated diagnostics and reports deployed for wider consumption.
  There are two problems with the current
  practice. First, there are gaps in this workflow: EDA is performed
  with one set of tools, and automated reports and deployments with another.
  Second, EDA environments often assume a single-developer perspective,
  while data scientist teams would get much benefit from easier sharing
  of scripts and data feeds, experiments, annotations, and automated
  recommendations, which are well beyond what traditional version control
  systems provide. \todo{sell idea, not system} In this paper, we propose RCloud, a system that supports
  collaborative data analysis, visualization and web deployment. We will
  discuss the design decisions, tradeoffs and limitations, and compare RCloud
  to other current proposals.
} % end of abstract

>>>>

\title{RCloud: Integrating Exploratory Visualization, Analysis and Deployment}
%% \title{RCloud: DevOps for Visual Analysis and Data Science}

@@

%% Abstract section.
\abstract{ Consider the emerging role of a data science team within an
  organization. Individual data scientists and statisticians usually
  work on loosely related problems, and must devise effective ways to
  share their findings and move results from exploratory data analysis
  into automated diagnostics and reports deployed for wider consumption.
  There are two problems with the current
  practice. First, there are gaps in this workflow: EDA is performed
  with one set of tools, and automated reports and deployments with another.
  Second, EDA environments often assume a single-developer perspective,
  while data scientist teams would get much benefit from easier sharing
  of scripts and data feeds, experiments, annotations, and automated
  recommendations, which are well beyond what traditional version control
  systems provide. We contribute and justify the following three
  requirements for systems built to support current data science teams
  and users: \emph{technology transfer}, \emph{coexistence}, and
  \emph{discoverability}. In addition, we contribute the design and
  implementation of RCloud, a prototype system that supports the requirements
  collaborative data analysis, visualization and web deployment. We
  discuss the design decisions, tradeoffs and limitations, and compare RCloud
  to other current proposals.
} % end of abstract
